---
format: html
---

```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

I downloaded GBIF records for iNaturalist and Cornell Lab of Ornithology from 2015 to 2025 at Natural History Museum of Los Angeles County and Rose Garden.

## QA raw GBIF data

```{r}
#| message: false
nhm_raw_df <-read_tsv('../data/raw/0025996-250711103210423.csv')
dim(nhm_raw_df)
```

```{r}
#| message: false
rose_raw_df <-read_tsv('../data/raw/0025992-250711103210423.csv')
dim(rose_raw_df)
```
get count of taxon ranks 


```{r}
table(c(nhm_raw_df$taxonRank, rose_raw_df$taxonRank))
```
```{r}
temp_nhm <- nhm_raw_df |> filter(institutionCode == 'CLO')
temp_rose <- rose_raw_df |> filter(institutionCode == 'CLO')

table(c(temp_nhm$taxonRank, temp_rose$taxonRank))
```

```{r}
temp_nhm <- nhm_raw_df |> filter(institutionCode == 'iNaturalist')
temp_rose <- rose_raw_df |> filter(institutionCode == 'iNaturalist')

table(c(temp_nhm$taxonRank, temp_rose$taxonRank))
```

Get date for newest observation

```{r}
 nhm_raw_df |> 
  filter(institutionCode == 'CLO') |> 
  select(eventDate) |> summary()
```
```{r}
 nhm_raw_df |> 
  filter(institutionCode == 'iNaturalist') |> 
  select(eventDate) |> summary()
```

iNaturalist provides coordinateUncertaintyInMeters, CLO does not. Filter out iNaturalist records with high coordinateUncertaintyInMeters.

View distribution of coordinateUncertaintyInMeters.

```{r}
temp_nhm <- nhm_raw_df |>
  filter(coordinateUncertaintyInMeters < 100)

ggplot(data = temp_nhm,  mapping = aes(x = coordinateUncertaintyInMeters)) +
  geom_histogram()
```

```{r}
temp_rose <- rose_raw_df |>
  filter(coordinateUncertaintyInMeters < 100)

ggplot(data = temp_rose,  mapping = aes(x = coordinateUncertaintyInMeters)) +
  geom_histogram()
```

Count records by institutionCode.

```{r}
table(nhm_raw_df$institutionCode)
```
```{r}
table(rose_raw_df$institutionCode)
```

## clean raw GBIF data

```{r}
clean_gbif <- function(df) {
  df |>
  filter((institutionCode == 'iNaturalist' & 
            coordinateUncertaintyInMeters < 100 ) | 
           (institutionCode == 'CLO'))  |>
  filter(year %in% c(2014:2023)) |>
  filter(taxonRank %in% c('FORM', 'SPECIES', 'SUBSPECIES', 'VARIETY'))  
}
```

```{r}
nhm_df <-  clean_gbif(nhm_raw_df) |>
  mutate(place='NHMLAC')

dim(nhm_df)
```

`
```{r}
rose_df <- clean_gbif(rose_raw_df) |>
  mutate(place='Rose Garden')

dim(rose_df)

```


Combine dataframes

```{r}
combine_df <- bind_rows( nhm_df, rose_df )
dim(combine_df)
```


```{r}
write_csv(combine_df, '../data/processed/gbif_gardens.csv', na='')
```
